\textbf{How open are hybrid journals included in transformative agreements?}

### 2024-02-15 analysis/manuscript.tex
minor

Throughout this mostly automated data gathering and analysis process, we
used tools from the Tidyverse (Wickham et al., 2019) for the R
programming language (R Core Team, 2020). The resulting data is openly
available through an R package, hoaddata. Following Marwick et al.
(2018), hoaddata contains not only the datasets used in the data
analysis. It also includes code used to compile the data by connecting
it the cloud-based Google Big Query data warehouse, where the big
scholarly data from Crossref, OpenAlex and Unpaywall were imported. To
increase the computational reproducibility, the R package was
automatically built using GitHub Actions, a continuous integration
service.

@@

\caption{Relative growth of open access in hybrid journals in transformative agreements between 2018 and 2022 per publication year. The blue areas represent open access through transformative agreements, the grey areas depict open access articles where no link to an agreement could be established (according to matching OpenAlex first author affiliations matched with cOAlition S transformative agreement data). (A) Proportion of open access articles in hybrid journals per year. (B) Percentage of hybrid open access via agreements per year. Boxplots show the proportion of open access articles by individual hybrid journals (C) and individual open access uptake rates by individual hybrid journals and open access funding (D) per publication year. The individual outliers are not shown. Note that data on transformative agreements ending before June 2021 were not available for this study.}\label{fig:results_overview}
\end{figure}

@@

@@

Analysing hybrid open access across publishers between 2018 and 2022
reveals a large market concentration. Although 48 publishers offered
transformative agreements, the big three commercial publishers Elsevier,
Springer Nature, and Wiley, accounted for 49\% of total article volume
published (see Table \ref{tab:publisher_league_table}). Together, they
published 500,878 open access articles, representing 66\% of the open
access articles in hybrid journals. Elsevier, Springer Nature, and Wiley
made 243,891 articles open access in hybrid journals through
transformative agreements, resulting in an even larger market share of
74\%. However, there are differences among the three large publishers.
Although Elsevier published the largest volume of articles (n =
2,770,826, 34\%), it published a relatively low number of open access
articles, including those that can be associated with transformative
agreements. In contrast, Springer Nature and Wiley provided open access
to a larger proportion of their articles (13\% of Springer Nature
articles and 15\% of Wiley articles were open acccess), leading to
higher open access market shares (23\% Springer Nature resp. 23\%
Wiley). This difference between Elsevier on the one hand and Springer
Nature and Wiley on the other can be attributed to transformative
agreements, as the latter made the majority of their open access
articles available through such deals (Springer Nature 57\% resp. Wiley
55\%).

@@

\caption{Developement of open access in hybrid journals included in transformative agreements between 2018 and 2022 by publishers. The blue areas represent open access through transformative agreements, the grey areas depict open access articles where no link to an agreement could be established (according to matching OpenAlex first author affiliations matched with cOAlition S transformative agreement data). (A) Proportion of open access articles in hybrid journals per year and publisher. (B) Percentage of hybrid open access via agreements per year and publisher. Boxplots (C) show individual open access uptake rates by individual hybrid journals and open access funding per publication year and publisher. The individual outliers are not shown. Note that data on transformative agreements ending before June 2021 were not available for this study.}\label{fig:publisher_figure}
\end{figure}

The varying degrees of uptake of open access across the three major
publishers can be attributed to distinct approaches to transformative
agreements. Springer Nature, for example, began in 2015 offering
selected consortia, such as the Max Planck Society, the Swedish Bibsam
consortium, and the Finnish FinELib consortium, open access agreements
for its hybrid journal portfolio under the name Springer
Compact\footnote{\url{https://web.archive.org/web/20180414062853id_/http://www.liber2015.org.uk/wp-content/uploads/2015/03/Springer-Compact.pdf}}.
However, these agreements were not included in the data as they
concluded prior to the start of the transformative agreement data
collection in June 2021. Nonetheless, the results suggest the importance
of central agreements for Springer Nature's hybrid open access business
over the past five years (Figure 2B). In 2022, 66\% (n = 34,725) of open
access articles in f Springer Nature hybrid journals were enabled
through transformative agreements. In the same year, 70\% (n = 37,316)
of Wiley's open access articles could be linked to transformative
agreements in 2022. In contrast, Elsevier published fewer than half of
its open access articles through transformative agreements (n = 32,627;
54\%).

The increasing trend towards transformative agreements can be also
observed at the journal-level (Figure \ref{fig:publisher_figure}C).
While no substantial differences between open access enabled through
transformative agreements and other revenue source could observed across
Elsevier journals, the distribution of open access across Springer
Nature and Wiley hybrid journals indicates that the growth is not
limited to a few journals, but extends across the portfolio. In
particular, Wiley's upper quantile, which represents the top 25\% of
journals in terms of the proportion of open access articles from
transformative agreements, increased markedly from 13\% in 2020 to 31\%
in 2022. At the same time, the median proportion grew from 7.5\% to
19\%. It is interesting to note that a small but increasing number of
journals from these two publishers are providing open access to the
majority of articles through transformative agreements. Wiley recorded
68 and Springer Nature 102 hybrid journals with an open access share
above 50\% that could be solely attributed to transformative agreements.
Upon inspection, these journals were mainly society or local language
journals with a small yearly article volume.

@@

Figure \ref{fig:subject_panel} presents the relative growth of hybrid
open access by subject area between 2018-2022. In particular, Social
Sciences and Humanties journals accounted for the strongest growth in
the five-years period from 6.4\% (n = 8,361) to 23\% (n = 51,938),
followed by the Life Science from 7.6\% (n = 15,003) to 18\% (n =
39,494) , Health Science from 5.3\% (n = 18,279) to 16\% (n = 63,089)
and Physical Sciences from 4.5\% (n = 22,364) to 12\% (n = 85,428). This
growth in the social sciences can be largely attributed to
transformative agreements. In 2022, two-third of open access articles
(67\%, n = 34,759) were published by first authors affiliated with
participating institutions (see \ref{fig:subject_panel}B). Figure
\ref{fig:subject_panel}C shows that this trend is consistent across
Social Sciences journals. In 2022, 25\% of Social Science journals
provided open access to at least every fourth article exclusively
through transformative agreements. However, hybrid open access through
transformative agreements played a comparable lesser role in the Life
Sciences and Health Sciences. In these two subject areas, only about
half of the open access articles can be linked to these agreements, both
overall and on median average across journals. In contrast, the majority
of Physical Science Journals, shows an increase of open access through
transformative agreements compared to other options to publish open
access in hybrid journals.

\begin{center}\includegraphics[width=0.99\linewidth,]{fig/subject_panel-1} \end{center}

@@

Between 2018 and 2022, Western economies almost exclusively dominated
hybrid open access publishing through transformative agreements. During
this period, first-authors affiliated with institutions from
Organisation for Economic Co-operation and Development (OECD) member
countries published 602,050 open access articles in hybrid journals,
representing 81\% of the investigated open access articles. This
disparity between OECD nations and other countries becomes even more
evident when considering open access through transformative agreements,
as 310,712 of 328,957, or 94\% of open access articles were associated
with such agreements.

@@

\caption{Development of hybrid open access publishing by countries}\label{fig:country_patch}
\end{figure}

Figure \ref{fig:country_patch}A shows the development of hybrid open
access publishing by countries, comparing the OECD area with the BRICS,
an intergovernmental organisation, which comprised the countries Brazil,
Russia, India, China and South Africa in 2022. The residual category
``Other'' includes the remaining countries. From 2018 to 2022, the
proportion of open access in hybrid journals increased from 6.1\% in
2018 to 26\% in 2022. On the other hand, BRICS recorded an low uptake,
from 1.6\% in 2018 to 3.7\% in 2022.\\
Despite rise of open access across OECD countries, the overall
publication output decreased sharply, dropping to 786,903 in 2022 after
peaking 892,197 articles in 2020. In stark contrast, the number of
articles published in hybrid journals by first authors affiliated with
institutions from BRICS countries increased steadily over the years,
more than doubling from 356,632 in 2018 to 786,903 in 2022. Upon closer
examination, this trend can be observed across all big three publishers,
although the shift towards BRICS is particularly evident in Elsevier's
hybrid journal portfolio, in particular with regard to articles
published in Physical Sciences journals (see Supplement). While OECD
publication output in Elsevier's Physical Sciences journals declined
from 112,822 articles in 2018 to 103,766 in 2022, BRICS output increased
from 104,654 to 171,713 in the same five-year period. Furthermore, OECD
publication output in Health Science Journals and Life Science journals
stagnated across the investigated hybrid journal portfolios after a peak
in 2020.

@@

\caption{Top 20 countries}\label{fig:country_top_20_plot}
\end{figure}

@@

\hypertarget{supplements}{%
\section{Supplements}\label{supplements}}

\begin{center}\includegraphics[width=0.99\linewidth,]{fig/unnamed-chunk-7-1} \end{center}

>>>>

Throughout this mostly automated data gathering and analysis process,
tools from the Tidyverse (Wickham et al., 2019) for the R programming
language (R Core Team, 2020) were used. The resulting data is openly
available through an R package, hoaddata, version 0.2.91 Following
Marwick et al. (2018), hoaddata contains not only the datasets used in
the data analysis. It also includes code used to compile the data by
connecting it the cloud-based Google Big Query data warehouse, where the
big scholarly data from Crossref, OpenAlex and Unpaywall were imported.
To increase the computational reproducibility, the R package was
automatically built using GitHub Actions, a continuous integration
service.

@@

\caption{Relative growth of open access in hybrid journals in transformative agreements between 2018 and 2022 per publication year. The blue areas represent open access through transformative agreements, the grey areas depict open access articles where no link to an agreement could be established. (A) Proportion of open access articles in hybrid journals per year. (B) Percentage of hybrid open access via agreements per year. Boxplots show the proportion of open access articles by individual hybrid journals (C) and individual open access uptake rates by individual hybrid journals and open access funding (D) per publication year. The individual outliers are not shown. Note that data on transformative agreements ending before June 2021 were not available for this study.}\label{fig:results_overview}
\end{figure}

@@

Analysing hybrid open access across publishers between 2018 and 2022
reveals a large market concentration. Although 48 publishers offered
transformative agreements, the big three commercial publishers Elsevier,
Springer Nature, and Wiley, accounted for 49\% of total article volume
published (see Table \ref{tab:publisher_league_table}). Together, they
published 500,878 open access articles, representing 66\% of the open
access articles in hybrid journals. Elsevier, Springer Nature, and Wiley
made 243,891 articles open access in hybrid journals through
transformative agreements, resulting in an even larger market share of
74\%.

@@

However, there are notable differences among the three large publishers.
Although Elsevier published the largest volume of articles (n =
2,770,826, 34\%), it published a relatively low number of open access
articles, including those that can be associated with transformative
agreements. In contrast, Springer Nature and Wiley provided open access
to a larger proportion of their articles (13\% of Springer Nature
articles and 15\% of Wiley articles were open acccess), leading to
higher open access market shares (23\% Springer Nature resp. 23\%
Wiley). This difference between Elsevier on the one hand and Springer
Nature and Wiley on the other can be attributed to transformative
agreements, as the latter made the majority of their open access
articles available through such deals (Springer Nature 57\% resp. Wiley
55\%).

@@

\caption{Developement of open access in hybrid journals included in transformative agreements between 2018 and 2022 by publishers. The blue areas represent open access through transformative agreements, the grey areas depict open access articles where no link to an agreement could be established. (A) Proportion of open access articles in hybrid journals per year and publisher. (B) Percentage of hybrid open access via agreements per year and publisher. Boxplots (C) show individual open access uptake rates by individual hybrid journals and open access funding per publication year and publisher. The individual outliers are not shown. Note that data on transformative agreements ending before June 2021 were not available for this study.}\label{fig:publisher_figure}
\end{figure}

The varying degrees of uptake of open access across the three major
publishers can be attributed to distinct approaches to transformative
agreements. Springer Nature, for example, began in 2015 offering
selected consortia, such as the Max Planck Society, the Swedish Bibsam
consortium, and the Finnish FinELib consortium, open access agreements
for its hybrid journal portfolio under the name Springer
Compact\footnote{\url{https://web.archive.org/web/20180414062853id_/http://www.liber2015.org.uk/wp-content/uploads/2015/03/Springer-Compact.pdf}}.
However, these agreements were not included in the data as they
concluded prior to the start of the transformative agreement data
collection in June 2021. Nonetheless, the results suggest the importance
of central agreements for Springer Nature's hybrid open access business
over the past five years (Figure 2B). In 2022, 66\% (n = 34,725) of open
access articles in Springer Nature hybrid journals were enabled through
transformative agreements. In the same year, 70\% (n = 37,316) of
Wiley's open access articles could be linked to transformative
agreements in 2022. In contrast, Elsevier published fewer than half of
its open access articles through transformative agreements (n = 32,627;
54\%).

The increasing trend towards transformative agreements can be also
observed at the journal-level (Figure \ref{fig:publisher_figure}C).
While no substantial differences between open access enabled through
transformative agreements and other revenue sources could observed
across Elsevier journals, the distribution of open access across
Springer Nature and Wiley hybrid journals indicates that the growth is
not limited to a few journals, but extends across the portfolio. In
particular, Wiley's upper quantile, which represents the top 25\% of
journals in terms of the proportion of open access articles from
transformative agreements, increased markedly from 13\% in 2020 to 31\%
in 2022. At the same time, the median proportion grew from 7.5\% to
19\%. It is interesting to note that a small but increasing number of
journals from these two publishers are providing open access to the
majority of articles through transformative agreements. Wiley recorded
68 and Springer Nature 102 hybrid journals with an open access share
above 50\% that could be solely attributed to transformative agreements.
Upon inspection, these journals were mainly society or local language
journals with a small yearly article volume.

@@

Figure \ref{fig:subject_panel} presents the relative growth of hybrid
open access by subject area between 2018-2022. In particular, Social
Sciences and Humanties journals accounted for the strongest growth in
the five-years period from 6.4\% (n = 8,361) to 23\% (n = 51,938),
followed by the Life Science from 7.6\% (n = 15,003) to 18\% (n =
39,494) , Health Science from 5.3\% (n = 18,279) to 16\% (n = 63,089)
and Physical Sciences from 4.5\% (n = 22,364) to 12\% (n = 85,428). This
growth in the social sciences can be largely attributed to
transformative agreements. In 2022, two-third of open access articles
(67\%, n = 34,759) were published by first authors affiliated with
participating institutions (see \ref{fig:subject_panel}B).

{\centering \includegraphics[width=0.99\linewidth,]{fig/subject_panel-1}

\caption{Developement of open access in hybrid journals in transformative agreements between 2018 and 2022 by AJCS subject area. The blue areas represent open access through transformative agreements, the grey areas depict open access articles where no link to an agreement could be established. (A) Proportion of open access articles in hybrid journals per year and publisher. (B) Percentage of hybrid open access via agreements per year and publisher. Boxplots (C) show individual open access uptake rates by individual hybrid journals and open access funding per publication year and publisher. The individual outliers are not shown. Note that data on transformative agreements ending before June 2021 were not available for this study.}\label{fig:subject_panel}
\end{figure}

Figure \ref{fig:subject_panel}C shows that this trend is consistent
across Social Sciences journals. In 2022, 25\% of Social Science
journals provided open access to at least every fourth article
exclusively through transformative agreements. However, hybrid open
access through transformative agreements played a comparable lesser role
in the Life Sciences and Health Sciences. In these two subject areas,
only about half of the open access articles can be linked to these
agreements, both overall and on median average across journals. In
contrast, the majority of Physical Science Journals, shows an increase
of open access through transformative agreements compared to other
options to publish open access in hybrid journals.

@@

Between 2018 and 2022, high-income economies almost exclusively
dominated hybrid open access publishing through transformative
agreements. During this period, first-authors affiliated with
institutions from Organisation for Economic Co-operation and Development
(OECD) member countries published 602,050 open access articles in hybrid
journals, representing 81\% of the investigated open access articles.
This disparity between OECD nations and other countries becomes even
more evident when considering open access through transformative
agreements, as 310,712 of 328,957, or 94\% of open access articles were
associated with such agreements.

Figure \ref{fig:country_patch}A shows the development of hybrid open
access publishing by countries, comparing the OECD area with the BRICS,
an intergovernmental organisation, which comprised the countries Brazil,
Russia, India, China and South Africa in 2022. The residual category
``Other'' includes the remaining countries. From 2018 to 2022, the
proportion of open access in hybrid journals increased from 6.1\% in
2018 to 26\% in 2022. On the other hand, BRICS recorded an low uptake,
from 1.6\% in 2018 to 3.7\% in 2022.\\
Despite rise of open access across OECD countries, the overall
publication output decreased sharply, dropping to 786,903 in 2022 after
peaking 892,197 articles in 2020. In stark contrast, the number of
articles published in hybrid journals by first authors affiliated with
institutions from BRICS countries increased steadily over the years,
more than doubling from 356,632 in 2018 to 786,903 in 2022. Upon closer
examination, this trend can be observed across all big three publishers,
although the shift towards BRICS is particularly evident in Elsevier's
hybrid journal portfolio, in particular with regard to articles
published in Physical Sciences journals. While OECD publication output
in Elsevier's Physical Sciences journals declined from 112,822 articles
in 2018 to 103,766 in 2022, BRICS output increased from 104,654 to
171,713 in the same five-year period. Furthermore, OECD publication
output in Health Science Journals and Life Science journals stagnated
across the investigated hybrid journal portfolios after a peak in 2020.

@@

\caption{Development of hybrid open access publishing by country. (A) presents the number of articles published in hybrid journals included in transformative agreement, distinguishing between BRICS as of 20222, OECD and other countries. (B) Scatterplot contrasting total articles with open access article volume in 2022, by country. Point size represents the number of articles enabled through transformative agreements. (C) Hybrid open access market share in 2022 by country. In (A) and (C), the blue areas represent open access through transformative agreements, the grey areas depict open access articles where no link to an agreement could be established. Country names are represented as ISO two-letter country codes.}\label{fig:country_patch}
\end{figure}

@@

\caption{Development of open access in hybrid journals in transformative agreements between 2018 and 2022, by the Top 20 most productive countries in terms of total articles published in the five-years period. Blue areas represent open access through transformative agreements, the grey areas depict open access articles where no link to an agreement could be established. Country names are represented as ISO two-letter country codes}\label{fig:country_top_20_plot}
\end{figure}

@@

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

\hypertarget{conclusions}{%
\section{Conclusions}\label{conclusions}}

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}
